---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gam2formula

<!-- badges: start -->
[![R-CMD-check](https://github.com/iqtigorg/gam2formula/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/iqtigorg/gam2formula/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The `gam2formula` package converts spline smooths from generalized additive models
fitted with `mgcv` into closed-form algebraic formulas that can be
printed, exported to LaTeX, and predicted from.

## Installation

You can install the current version of `gam2formula` with:

``` r
remotes::install_github("iqtigorg/gam2formula")
```

## Example

Fit a spline in a generalized additive model using `mgcv`:

```{r example0, message=FALSE, out.width="75%"}
library(mgcv)
m <- gam(mpg ~ s(qsec, bs = "cr", k = 5), data = mtcars)
plot(m)
```

Display the algebraic formula of a spline as coefficient table using `gam2formula` :

```{r example1}
library(gam2formula)
mod_formulas <- gam2formula(m)
print(mod_formulas, term = "qsec")
```

And use the formula for point predictions, independent of the original model object:

```{r example2, out.width="75%"}
plot(m)
points(15:22, predict(mod_formulas, term = "qsec", newdata = data.frame(qsec = 15:22)))
```

For further examples and documentation, see our [vignette](inst/doc/using-gam2formula.pdf).
