---
title: "Using `gam2formula`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
geometry: margin=0.8in
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Using gam2formula}
  %\VignetteEngine{knitr::rmarkdown}
---

\newpage

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

Load packages:

```{r, warning = FALSE}
library(mgcv)
library(gam2formula)
```

# Gaussian GAM

## Simulate data and fit model with `mgcv`

Fit a GAM with `mgcv` to simulated data using B-splines, P-splines and cubic regression splines, among others:
 
```{r, generate}
set.seed(2)
dat <- gamSim(1, n = 200, dist = "normal", scale = 2, verbose = FALSE)
mod <- gam(y ~ s(x0, bs = "cr") +
               s(x1, bs = "bs", m = c(4, 2)) +
               s(x2, bs = "ps") + 
               s(x3), data = dat)
```

Plot the smooth terms with `mgcv`'s internal plot-function:

```{r, plot}
plot(mod, pages = 1)
```

\newpage

## Derive closed formulas for all supported smooth terms

Applying `gam2formula()` derives closed formulas for all supported smooth terms. Use `summary()` to get an overview:

```{r, gam2formula}
mod_formulas <- gam2formula(mod)
summary(mod_formulas)
```

## Print derived formulas

To print a derived formula, use `print()` with the desired `term`.

### Cubic regression spline for `x0`

Let's look at the formula for the predictor `x0`, modeled with a cubic regression spline:

```{r, inspect1}
print(mod_formulas, term = "x0")
```

### B-spline for `x1`

Let's look at the formula for the predictor `x1`, modeled with a B-spline of degree 4:

```{r, inspect2}
print(mod_formulas, term = "x1")
```


### P-spline for `x2`

Let's look at the formula for the predictor `x2`, modeled with a P-spline of degree 3:

```{r, inspect3}
print(mod_formulas, term = "x2")
```

\newpage

## Compare to internal predictions by `mgcv`

As a quick visual check, we can make predictions for a vector of new predictor values based on our obtained formulas, using `predict()`, and compare it to the internal predictions plot from `mgcv`. We deliberately choose new values of the predictor partially outside of the observed range to also show the expected behavior for extrapolation (which is linear by `mgcv`'s default).
 
### Cubic regression spline for `x0`

```{r, compare1}
# generate new predictor data
x0_newmin <- min(dat$x0) - 0.2
x0_newmax <- max(dat$x0) + 0.2
x0_new <- seq(x0_newmin, x0_newmax, length.out = 30)

# compute the linear predictor from coefficient table for new data points
y <- predict(mod_formulas, term = "x0", newdata = data.frame(x0 = x0_new))

# plot comparison
plot(mod, select = 1, xlim = c(x0_newmin, x0_newmax))
points(x0_new, y, col = "dodgerblue", pch = 16)
```

### B-spline for `x1`

```{r, compare2}
# generate new predictor data
x1_newmin <- min(dat$x1) - 0.2
x1_newmax <- max(dat$x1) + 0.2
x1_new <- seq(x1_newmin, x1_newmax, length.out = 30)

# compute the linear predictor from coefficient table for new data points
y <- predict(mod_formulas, term = "x1", newdata = data.frame(x1 = x1_new))

# plot comparison
plot(mod, select = 2, xlim = c(x1_newmin, x1_newmax))
points(x1_new, y, col = "olivedrab", pch = 16)
```

### P-spline for `x2`

```{r, compare3}
# generate new predictor data
x2_newmin <- min(dat$x2) - 0.2
x2_newmax <- max(dat$x2) + 0.2
x2_new <- seq(x2_newmin, x2_newmax, length.out = 30)

# compute the linear predictor from coefficient table for new data points
y <- predict(mod_formulas, term = "x2", newdata = data.frame(x2 = x2_new))

# plot comparison
plot(mod, select = 3, xlim = c(x2_newmin, x2_newmax))
points(x2_new, y, col = "tomato", pch = 16)
```

\newpage

## Print formulas in LaTeX

We can also print the formulas in LaTeX using the `print()` function. The resulting LaTeX code can be pasted directly into a LaTeX editor of your choice. When using `results = 'asis'` for the chunk options, the LaTeX code will be compiled directly in an RMarkdown script like the current vignette.

### Cubic regression spline for `x0`

```{r, latexform1, results='asis'}
print(mod_formulas, term = "x0", format = "latex")
```

### B-spline for `x1`

```{r, latexform2, results = 'asis'}
print(mod_formulas, term = "x1", format = "latex")
```

### P-spline for `x2`

```{r, latexform3, results = 'asis'}
print(mod_formulas, term = "x2", format = "latex")
```

\newpage

# Bernoulli GAM

## Fit model with `mgcv`

As real example data, we use the original Pima Indians Diabetes data set as provided, for instance, in the `mlbench` package. We predict binary diabetes status using cubic regression splines for age and glucose level and plot the estimated effects.

```{r, fitbinary}
library(mlbench)
data(PimaIndiansDiabetes)
PimaIndiansDiabetes$diabetes <- as.numeric(PimaIndiansDiabetes$diabetes == "pos")
mod <- gam(diabetes ~ s(age, bs = "cr", k = 8) + s(glucose, bs = "cr", k = 8), 
         family = "binomial", data = PimaIndiansDiabetes)
plot(mod, pages = 1)
```

## Print formulas in LaTeX

```{r, printbinary, results = 'asis'}
mod_formulas <- gam2formula(mod)
print(mod_formulas, term = "age", format = "latex")
print(mod_formulas, term = "glucose", format = "latex")
```

## Handling point predictions

To get the predicted probability of diabetes for two newly observed individuals using `predict()`, we need to build the linear predictor from its three components (an intercept term, the age term, and the glucose term) and apply the inverse link function:

```{r, predbinary}
newdata <- data.frame(age = c(22, 53), glucose = c(92, 131)) # new observations

linpred <- coef(mod)["(Intercept)"] + # global intercept
  predict(mod_formulas, "age", newdata) + # linear predictor term for age
  predict(mod_formulas, "glucose", newdata) # linear predictor term for glucose

plogis(linpred)
```

The result is equivalent to the predicted probabilities from `mgcv::predict.gam()`:

```{r, predbinaryauto}
as.numeric(mgcv::predict.gam(mod, newdata, type = "response"))
```

Note that for models with many predictors, it can be convenient to use `mgcv`'s `exclude` argument to get all linear predictor contributions *except* the term of interest from `mgcv`, and then add the derived formula's prediction:

```{r, predbinaryexclude}
 # get full linear predictor except s(age) from mgcv
linpred <- as.numeric(predict(mod, newdata, exclude = "s(age)")) +
  predict(mod_formulas, "age", newdata) # then add age term gam2formula

plogis(linpred)
```
